\textbf{Master's Thesis} \\ Correct Projection Mapping on - and Interaction Techniques with - Cloth-based Deformable Surfaces

### 2014-08-18 MSc Report/MScThesis_DanAvram.tex
Fixed typos, reference errors, missing words and other minor things.

One of the main problems with deformable displays is that the image contents become distorted when deforming the surface - the image will follow the shape of the display. For many applications this is not desired: for example, if we want to move an object by pushing and dragging it around, the image should appear as on a flat screen - otherwise the object will also move around and become skewed, due to the deformation induced by the display's shape.

@@

In \cite{gokturk}, the authors describe the TOF technology, and state that devices that use it are enabled to perceived and interact with the world around them. Further, the technology is succinctly described as sending out a signal(in this case, light) and measuring a property of the signal(phase shift - differences between sinusoidal wave patterns) that returns from a target. This property can be used to obtain the time of flight. By multiplying it with the velocity of the signal, one can obtain the traveled distance. Lastly, they mention some issues with noise and state that their behavior depends on the amount of light reflected into the sensor.
\end{itemize}

@@

\textbf{The Deformable Workspace}(see figure \ref{fig:deformable_workspace}) is another approach to creating a deformable screen that acts as a boundary surface between the real and virtual worlds and is presented in [5]. The workspace is set up in similar ways to other techniques by using an IR camera, IR projector and an LCD projector. Basically, the IR camera and projector could be replaced by a depth sensor such as the Kinect.

@@

In [5] they also provide a method to compensate for projection distortion, given that both the calibration parameters and the shape of the deformed screen are known. The technique involves pre-warping the image by shifting each projected point. Further, they claim that PTM can be used to calculate and render the warped image to be projected in real-time, efficiently and easily. A limitation of the previous approach is that an assumption has been made that the user’s eyes are fixed. However, a viewpoint-based projection should be employed to accommodate realistic scenarios - the virtual space must be rotated and transformed according to the change of viewpoint.\\

@@

In order to get a better overview and understanding of the whole process required to correct the distortions, I have created a system pipeline(see figure \ref{fig:Pipeline}) that describes the whole process, at a high-level. Each part will be described individually in subsequent chapters.

@@

The procamlib (https://code.google.com/p/procamcalib/) software package for Matlab can be used to calibrate the projector-camera system using this approach. The steps are explained in \cite{falcao08}. However, both the projected checkerboard and the printed one need to be in view of the camera at the same time. This is not possible with the setup used in this project.\\
\item 3D Projection Mapping by Using a Reconstructed 3D Object\\

@@

The naive triangulation, by connecting the triangles without considering some measure of how natural it looks, may not provide the intuitively \textit{best-looking} results. If the triangulation is not acceptable, it can be improved with a technique such as \textbf{Delaunay Triangulation}(explained in detail in \cite[Chapter~9]{berg08}), where triangulations are ranked by comparing their smallest angle(small angles, result in \textit{skinniness}). This tecnhique can be used for 2.5D reconstruction and can obtain an optimal triangulation. For more information please consult appendix \ref{appendix:B-delaunay}.

@@

As described in section \ref{sec:TriangleStripTriangulation} a triangle strip has been used to triangulate the point cloud. In order to do so, an \textit{IBO(Index Buffer Object)} has been used to store a \verb|GL_ELEMENT_ARRAY_BUFFER|. This object is very similar to a VBO, the only difference is that we do not store vertex attributes in the element array, but actual element indices. In other words, we need to compute an array of indices, that defines how the triangle strip should be connected, and then just simply bind the data to the IBO:
\begin{verbatim}
ibo.addData(&(rdp->indices[0]), 
    rdp->indices.size() * sizeof(unsigned int));
ibo.uploadDataToGPU(GL_STATIC_DRAW);
\end{verbatim}

We add an array of \verb|int|s into the buffer object of the size of the depth camera's resolution, in pixels(\verb|width*height|). Note that the index computation can only be done once! We always have the same indices, for the same grid, since the x and y values are fixed for a vertex - the indices are computed for all points, even for invalid ones, this means we do not have to recompute the indices if the state of the points changes(since the VBO data for those invalid points is not initialized, they will not be drawn). This means the flag for data optimization has been set to \verb|GL_STATIC_DRAW|. Further, the \verb|uploadDataToGPU| method internally calls  \verb|glBufferData| with the previously added element array data. Finally, the IBO is bound and setup similarly to a VBO.

@@

The parameters are briefly explained in the comments, but to detail briefly:
\begin{itemize}
\item the first parameter is the type of primitive OpenGL will use for rendering
\item the second parameter represents the number of elements required to draw the mesh
\item the third parameter defines the type of the elements 
\item finally, the last parameter is simply a pointer that defines an offset, but since we only have the mesh data to render with indices, this can be set to 0
\end{itemize}

@@

Additionally, we must only apply the algorithm if all the neighbors of the current point have valid depth values(since certain areas of the image do not contain depth information), so this must also be checked.

@@

We need to compute the texture coordinates for the projected texture and pass them to the fragment shader:
\begin{verbatim}
out vec4 projCoords;
vec4 posEye    = MV * vec4(inPosition, 1.0);
vec4 posWorld  = InvViewMat * posEye;
projCoords     = TexGenMat * posWorld;
\end{verbatim}

The \verb|TexGenMat| has been constructed as described in section \ref{sec:ptm}, in the object linear texgen equation, except the final multiplication with \verb|M|:
\begin{verbatim}
texGenMatrix = biasMatrix * projectorP * projectorV;
invViewMatrix = glm::inverse(mModelView);
\end{verbatim}

@@

Further more, the \verb|debug| flag was used later on in order to change the color of the vertices, if they are within a certain depth range.

@@

An application that allows a user to manually specify a number of points in the image and subsequently returns the depth values corresponding to these points, has been developed(see figure \ref{fig:PointDepthExtraction}). It is using the OpenCV(see \cite{opencv}) and Intel Perceptual Computing libraries. The former was used for displaying the image, reacting to mouse input and drawing the crosses, while the latter provided the depth frame data from which the image was created. The depth has been returned in millimeters, and for the points in figure \ref{fig:PointDepthExtraction}, the values were: 689 for top left, 685 for bottom left, 701 for top right, 702 for bottom right. From these values, it appears that there is a small vertical offset(1-4 mm), and a larger offset on the horizontal axis(7-11 mm). The test was repeated several times, and the general pattern was the same - larger horizontal offset and smaller vertical offset. However, the obtained values shifted overall with up to 2-3 centimeters, between different frames and points. While the relative offset patterns were largely the same, there were a few exceptions, when the vertical offset was also large on one side. However, this issue, along with that of the points being shifted, may be caused due to noisy data, as the points in those cases were very near to the top or bottom edge of the valid depth data(where we can also see some small dark spots).

@@

Due to the fact that the cloth mesh is rotated, the texture can not be aligned perfectly(in order to do this, proper calibration is required; see figure \ref{fig:MisalignmentArea}). Otherwise, a texture with 4 red corners and a grid has been used to test if the grid looks fairly straight and the corners roughly correspond to the edges of the cloth.

@@

As mentioned, the camera was moved around the scene until it roughly had the texture in view. The field of view and distance of the camera have been adjusted to that effect, but, it is important to note that there is an infinity of ways to combine these two values in order to roughly have the texture in view. This is because, we can always increase the field of view, and decrease the z distance, or vice versa. Now, the important issue here, is that the image will not look the same for all different combinations of values! For example, if we move the camera closer to the mesh in the virtual world while decreasing the FOV, we will see more aliasing(noise), but the distortion compensation is also more pronounced. Conversely, if we move the camera further away while increasing the FOV, we will have a more stable image, but the distortion effect is less pronounced. In order to get accurate results, one needs to calibrate the system in order to obtain proper parameters. However, it is still possible to set the value to acceptable levels by trial and error.

@@

Since the system has not been properly calibrated, I do not expect the compensation to perform in a consistent way in all areas of the display. The uncompensated method is expected to be more consistent(since the image is not compensated).
\item Preference\\

@@

 The \textit{Grid Scenario} was chosen because it is easier to see if the vertical and horizontal lines remain relatively straight after deformation, compared to a detailed scene. Further, it is also interesting to see if text can retain it's shape after deformation, allowing the user to still read the words in the deformed area - for this reason, the \textit{Text Scenario} was used. Finally, the \textit{Columns Scenario} consists of a few objects of interest, namely columns. When one deforms the display, these columns bend, meaning that a compensated image should make them look straight. Further more, the display can also be deformed in the middle, between the columns, that also causes bending if the image is not compensated. Since the columns are placed on the sides, it is a good way to test the same objects, by interacting in different areas of the screen. In addition, the scene is 3D, so unlike the other scenarios, we can check if the depth between the objects is corrected.

 Due to these three scenarios being very similar, they will all be explained in one section. They all consisted of displaying a static image: of a grid, some text, and a number of columns in a scene, respectively. For each of these scenarios the users had to share their thoughts on methods A and B, for three different points in the image(see figure \ref{fig:GridTextColumns}). For these scenarios, they only used one finger interactions to deform the display at the indicated points(by using the index finger). Only one point was shown at a time, and it was made invisible after the participant remembered where the point was positioned.

@@

The map scenario simply consisted of a static map texture of a region of \textit{New York City City}(see figure \ref{fig:Map}). Users were asked to push into the display, using two hands. Further, they were also instructed to focus on the central area of the display. Then, the users were asked to compare the proportions and readability of the image, with both methods. Further, users were also instructed to focus on some particular name or landmark to focus on, as this makes it easier to compare.

@@

Since we are using a deformable display, one of the visualizations that makes sense to test is that of 3D volumes, represented using \textit{slices}(textures). This is because these slices are changed based on depth and we can simulate this by pushing an pulling into the display. Further, volume slicing interaction means we can also test how the compensation works when we have a dynamic scene/object. However, an issue with this scenario was that the slices were changing too fast, which in turn changed the shape of the object(brain) and made it more difficult for users to judge the compensation effect.

@@

One attempt to solve the problem was to use depth smoothing as described in section \ref{sec:DepthSmoothing}. The final result after spatial and temporal averaging was still not good enough - you could still see the difference between the two methods, because of the dynamic noise in the compensated image. However, the smoothing did reduce the overall noise and improved image quality.

@@

Some questions were also asked during the tasks. The questions are in the \textit{scenario notes} document, available under appendix \ref{appendix:C}. For the scenarios with the grid, text and columns feedback was collected regarding all of the three points tested. Similar scenario-specific questions were asked, as mentioned in the document(e.g.: how straight do you think the grid lines appeared with method B?).

@@

However, it is useful to also look in a bit more detail at statistical significance. The output of the t-tests is shown in table \ref{table:POA_Test}. Note that a \verb|*| symbol has been used to denote statistical significance. For all other cases, the null hypothesis could not be rejected. Since the confidence interval is 95\%, we would normally need a significance level of $<= 0.05$. However, we have a case of multiple testing, because we have one hypothesis tested for each scenario. As described in \cite{bonferroni}, we need to adjust the parameter $\alpha$, by dividing it with the number of tests(Bonferroni correction). In this case we have $\alpha = 0.05$ and $n$ = 5, which gives $\frac{\alpha}{n} = \frac{0.05}{5} = 0.01$. We must test against this value rather than the original $0.05$. However, in this case, the results are not affected.

@@

Moreover, the t-test results are shown in table \ref{table:OPOA_Test}; note that this time we only have one pair. As mentioned before, the missing values were replaced with 0 when running the t-test(the user was forced to select a preference for either method A or B). The results are statistically significant, with a value comfortably below the $0.05$ threshold(there is no need for a correction in this case, since the question is not asked for each scenario, so we only have one test).

@@

@@

Although there was no specific question regarding latency, several users mentioned that the compensation has a "time delay" or that "it takes some time to adjust"(there are 5 such occurrences in total, mentioned by 3 different participants). During the study, the scenarios were run at about 30 fps, because it had the depth smoothing functions enabled. However, this issue was also mentioned by the participant in the second pilot study, when the frame rate was about 60 fps. Therefore, this issue may not be related directly to rendering, although this issue has not been investigated.

@@

For this point, participants thought that the compensated method was preserving the appearance much  more often than the one which was not compensated(27 vs 3 occurrences). Some examples of items included in this category include: "keeps proportions better", "keeps the size of the object" for the uncompensated method and
"maintains proportions better", "better for the center", "resembles original better", "flattened image", "straighter". As we can see, the qualitative information is overwhelmingly in favor of the compensated method for this point.

@@

Regarding distortion, there were 19 comments for the compensated method ("distorts a lot", "deforming the image on the side", "the corner is more distorted", "it's going the wrong way - left, "letters go to the left", "moves to the left", "makes columns bend the other way") and only 6 for the uncompensated ("shifting everything", "doesn't compensate", "complete distortion", "perspective deformed"). From these comments we again see that the compensation is too much to the left of the deformed area, and actually, there are 8 such occurrences. This suggests this is the main problem, but one user also shared afterwards that "it was more corrected on the point", even though it looked very distorted. This happens because the setup is not calibrated, and the depth camera is not 100\% orthogonal to the display. This introduces misalignments near the sides if the camera is tilted, even slightly(which means the compensation will take place in a wrong area – with an offset; several users also complained of distortions “underneath the finger”).

@@

For this point, the preservation of appearance category has 14 entries\\("straighter", "preserves better", "fixing things around the finger", "maintains relative positions"), while only 3 for the uncompensated("more like the left side; keeps proportions better", "more straight" etc.). It is important to note, that although there are many more statements indicating that the compensating method preserves appearance, several of these also include additions like "not as good as center" or "better, but still bent", "not that straight as before(center)".

@@

In the comparison in figure \ref{fig:BrainScanComparison} we can see that the uncompensated method makes the object smaller as we push in, while the compensated one tries to retain its size. However, there may again be some overcompensation, since the image gets a bit larger than the original. A further issue, that may not be apparent in the video or the mentioned figure, is that overcompensation can also happen sideways, and then it seems that the shape of the object is being changed. However, both these issues are present(again) due to calibration issues(there is an offset towards the sides, and the amount of compensation changes with the field of view and distance of the virtual camera, as explained in section \ref{sec:VirtualCameraAlignmentToClothArea}.

@@

As described in section \ref{sec:Empirical Study Results}, we have obtained statistically significant quantitative data only for the preservation of appearance of the map scenario and the overall case, as well as the preference in the overall case. Even though the other scenario tests were not significant, the compensated method always scored higher. The results for the map scenario had the highest rating contrast between the methods. This is because the method which was not compensated made the text and roads very small and virtually unreadable, while pushing into the display. On the other hand, the compensated method retained the proportions fairly well.

Even though the consistency tests were not statistically significant, we could see that the results indicated that the behavior of the non-compensated method was more consistent. From the qualitative analysis presented in section \ref{sec:QualitativeAnalysis}, we can obtain more information about what differences participants perceived, and how they differ between the center, left and right points. The analysis indicates that the method compensated best for the center point, and worst for the left point. This issue is due to distortions that are introduced in the image by the compensated method(participants described this as a different kind of distortion), due to an offset compensation. This happens because the system is not calibrated and the components are not perfectly aligned. Although there were issues on the sides of the image, participants still selected the compensated method over the normal one, when asked to rate the methods over all cases. Since, as mentioned before, these tests are statistically significant, it is reasonable to believe that the compensation both preserved the image proportions better and was the preferred method for working with a deformable display.

@@

Moreover, an important extension is to provide a useful interface for developing interactive scenarios. Currently, only the brain scan scenario is interactive. Even so, it only checks for the depth of the center point in the image when changing textures, so it is a limited and naive approach. Although interaction is not the focus of this project, it is a necessary extension in order to thoroughly test the distortion correction in dynamic environments. For example, let's say we want to move a cube around in a 3D scene. In this case, latency becomes very important. It would be interesting to test if the cube would appear undistorted to users in such an environment.

@@

Deformable displays are a newly emerging type of natural user interfaces. One of the main issues related to such displays is the distortions that appear from the deformation of the object. The image will follow the shape of the displayed - it will not look like the intended output on a flat screen; this is called distortion.

@@

The compensation was implemented using a time-of-flight depth sensor, a standard projector and a cloth material(in a wooden frame) to display the images on. The model of the cloth is updated dynamically in a virtual scene(in real-time, up to about 60 fps, when not using noise reduction functions), and we can project any texture we want, to represent our final image contents. The depth camera is set to capture the back side of the display, on which we also project the image. When interacting from the front, pushing into the display is therefore modeled from the back-side, creating an extrusion in the corresponding area(see the thesis video for a demonstration). This enlarges the respective area, which negates the shrinking effect induced by moving that part of the display closer to the project - this negates the distortion, to a certain extent. The main components have therefore been tested and working, however, the system does have a few limitations, as described in the following paragraph.

@@

However, for a dense set of points, such that obtained from the depth data, a naive triangle strip triangulation has provided adequate results - the deformation is clearly visible, both in the mesh and on the texture applied to the mesh. For this reason, Delaunay triangulation has not been attempted, although it may improve visual quality. It remains to be determined, as future work, if the computational cost involved is worth the possible increase in visual quality.

@@

Many aspects from a classical implementation of marching cubes can also be used for a GPU implementation. This is the case with the basic look-up tables used to store edges and triangles. A version of these tables used in several implementations is available in an on-line article(see \cite{bourke1994}), where the full source code is also available. Although, there is no version implemented in a programmable shader on the GPU available, an implementation using a geometry shader is described in another on-line article(see \cite{crassin2007}). A geometry shader is simply a shader program that can be used to manipulate geometry directly - it allows the GPU to create geometry. This shader takes a single primitive as input(e.g.: a vertex) and outputs zero or more primitives(e.g.: zero or more triangles). 
This implementation is based on \cite{bourke1994}, with the main marching cubes logic inside the geometry shader. However, this implementation also uses deprecated OpenGL functionality. The main idea is that data can be stored inside 2D and 3D textures and can then be passed to the geometry shader as uniforms:

>>>>

One problem with deformable displays is that the image contents become distorted when deforming the surface - the image will follow the shape of the display. For many applications this is not desired: for example, if we want to move an object by pushing and dragging it around, the image should appear as on a flat screen - otherwise the object will also move around and become skewed, due to the deformation induced by the display's shape.

@@

In \cite{gokturk}, the authors describe the TOF technology, and state that devices that use it are enabled to perceive and interact with the world around them. Further, the technology is succinctly described as sending out a signal(in this case, light) and measuring a property of the signal(phase shift - differences between sinusoidal wave patterns) that returns from a target. This property can be used to obtain the time of flight. By multiplying it with the velocity of the signal, one can obtain the traveled distance. Lastly, they mention some issues with noise and state that their behavior depends on the amount of light reflected into the sensor.
\end{itemize}

@@

\textbf{The Deformable Workspace}(see figure \ref{fig:deformable_workspace}) is another approach to creating a deformable screen that acts as a boundary surface between the real and virtual worlds and is presented in \cite{watanabe08}. The workspace is set up in similar ways to other techniques by using an IR camera, IR projector and an LCD projector. Basically, the IR camera and projector could be replaced by a depth sensor such as the Kinect.

@@

In \cite{watanabe08} they also provide a method to compensate for projection distortion, given that both the calibration parameters and the shape of the deformed screen are known. The technique involves pre-warping the image by shifting each projected point. Further, they claim that PTM can be used to calculate and render the warped image to be projected in real-time, efficiently and easily. A limitation of the previous approach is that an assumption has been made that the user’s eyes are fixed. However, a viewpoint-based projection should be employed to accommodate realistic scenarios - the virtual space must be rotated and transformed according to the change of viewpoint.\\

@@

In order to get a better overview and understanding of the whole process required to correct the distortions, I have created a system pipeline(see figure \ref{fig:Pipeline}) that describes the whole process, at a high-level. Each part will be described individually in subsequent sections.

@@

The procamlib software package for Matlab(available at \cite{procamlib}) can be used to calibrate the projector-camera system using this approach. The steps are explained in \cite{falcao08}. However, both the projected checkerboard and the printed one need to be in view of the camera at the same time. This is not possible with the setup used in this project.\\
\item 3D Projection Mapping by Using a Reconstructed 3D Object\\

@@

The naive triangulation, by connecting the triangles without considering some measure of how natural it looks, may not provide the intuitively \textit{best-looking} results. If the triangulation is not acceptable, it can be improved with a technique such as \textbf{Delaunay Triangulation}(explained in detail in \cite[Chapter~9]{berg08}), where triangulations are ranked by comparing their smallest angle(small angles, result in \textit{skinniness}). This technique can be used for 2.5D reconstruction and can obtain an optimal triangulation. For more information please consult appendix \ref{appendix:B-delaunay}.

@@

As described in section \ref{sec:TriangleStripTriangulation} a triangle strip has been used to triangulate the point cloud. In order to do so, an \textit{IBO(Index Buffer Object)} has been used to store a \verb|GL_ELEMENT_ARRAY_BUFFER|. This object is very similar to a VBO, the only difference is that we do not store vertex attributes in the element array, but actual element indices. In other words, we need to compute an array of indices, that defines how the triangle strip should be connected, and then, bind the data to the IBO:
\begin{verbatim}
ibo.addData(&(rdp->indices[0]), 
    rdp->indices.size() * sizeof(unsigned int));
ibo.uploadDataToGPU(GL_STATIC_DRAW);
\end{verbatim}

We add an array of \verb|int|s into the buffer object of the size of the depth camera's resolution, in pixels(\verb|width*height|). Note that the index computation can be done only once! We always have the same indices, for the same grid, since the x and y values are fixed for a vertex - the indices are computed for all points, even for invalid ones, this means we do not have to recompute the indices if the state of the points changes(since the VBO data for those invalid points is not initialized, they will not be drawn). This means the flag for data optimization has been set to \verb|GL_STATIC_DRAW|. Further, the \verb|uploadDataToGPU| method internally calls  \verb|glBufferData| with the previously added element array data. Finally, the IBO is bound and setup similarly to a VBO.

@@

The parameters are briefly explained below:
\begin{itemize}
\item the first parameter is the type of primitive OpenGL will use for rendering
\item the second parameter represents the number of elements required to draw the mesh
\item the third parameter defines the type of the elements 
\item finally, the last parameter is simply a pointer that defines an offset, but since we only have the mesh data to render with indices, this can be set to 0
\end{itemize}

@@

Additionally, we must only apply the algorithm if all the neighbors of the current point have valid depth values(since certain areas of the image do not contain depth information).

@@

We need to compute the coordinates for the projected texture and pass them to the fragment shader. The \verb|TexGenMat| has been constructed as described in section \ref{sec:ptm}, in the object linear texgen equation, except the final multiplication with \verb|M|:
\begin{verbatim}
texGenMatrix = biasMatrix * projectorP * projectorV;
invViewMatrix = glm::inverse(mModelView);
\end{verbatim}

@@

@@

An application that allows a user to manually specify a number of points in the image and subsequently returns the depth values corresponding to these points, has been developed(see figure \ref{fig:PointDepthExtraction}). It is using the OpenCV(see \cite{opencv}) and Intel Perceptual Computing libraries. The former was used for displaying the image, reacting to mouse input and drawing the crosses, while the latter provided the depth frame data from which the image was created. The depth has been returned in millimeters, and for the points in figure \ref{fig:PointDepthExtraction}, the values were: 689 for top left, 685 for bottom left, 701 for top right, 702 for bottom right. From these values, it appears that there is a small vertical offset(1-4 mm), and a larger offset on the horizontal axis(7-11 mm). The test was repeated several times, and the general pattern was the same - larger horizontal offset and smaller vertical offset. However, the obtained values shifted overall with up to 2-3 centimeters, between different frames and points. While the relative offset patterns were largely the same, there were a few exceptions, when the vertical offset was also large on one side. This issue, along with that of the points being shifted, may be caused due to noisy data, as the points in those cases were very near to the top or bottom edge of the valid depth data(where we can also see some small dark spots).

@@

Due to the fact that the cloth mesh is rotated, the texture can not be aligned perfectly(in order to do this, proper calibration is required; see figure \ref{fig:MisalignmentArea}). A texture with 4 red corners and a grid, has been used to test how straight the grid looks and how well the corners correspond to the edges of the cloth.

@@

As mentioned, the camera was moved around the scene until it roughly had the texture in view. The field of view and distance of the camera have been adjusted to that effect, but, it is important to note that there is an infinity of ways to combine these two values in order to roughly have the texture in view. This is because, we can always increase the field of view, and decrease the z distance, or vice versa. Now, the important issue here, is that the image will not look the same for all different combinations of values! For example, if we move the camera closer to the mesh in the virtual world while decreasing the FOV, we will see more aliasing(noise), but the distortion compensation is also more pronounced. Conversely, if we move the camera further away while increasing the FOV, we will have a more stable image, but the distortion effect is less pronounced. In order to get accurate results, one needs to calibrate the system in order to obtain proper parameters. However, it is still possible to set the values to acceptable levels by trial and error.

@@

Since the system has not been properly calibrated, I do not expect the compensation to perform in a consistent way in all areas of the display. The uncompensated method is expected to be more consistent.
\item Preference\\

@@

 These scenarios all consisted of displaying a static image: of a grid, some text, and a number of columns in a scene, respectively. The \textit{Grid Scenario} was chosen because it is easier to see if the vertical and horizontal lines remain relatively straight after deformation, compared to a detailed scene. Further, it is also interesting to see if text can retain it's shape after deformation, allowing the user to still read the words in the deformed area - for this reason, the \textit{Text Scenario} was used. Finally, the \textit{Columns Scenario} consists of a few objects of interest, namely columns. When one deforms the display, these columns bend, meaning that a compensated image should make them look straight. Further more, the display can also be deformed in the middle, between the columns, that also causes bending if the image is not compensated. Since the columns are placed on the sides, it is a good way to test the same objects, by interacting in different areas of the screen. In addition, the scene is 3D, so unlike the other scenarios, we can check if the depth between the objects is corrected.

 For each of these scenarios the users had to share their thoughts on methods A and B, for three different points in the image(see figure \ref{fig:GridTextColumns}). For these scenarios, they only used one finger interactions to deform the display at the indicated points(by using the index finger). Only one point was shown at a time, and it was made invisible after the participant remembered where the point was positioned.

@@

The map scenario simply consisted of a static map texture of a region of \textit{New York City City}(see figure \ref{fig:Map}). Users were asked to push into the display, using two hands. Further, they were also instructed to focus on the central area of the display. Then, the users were asked to compare the proportions and readability of the image, with both methods. Further, users were also instructed to focus on some particular name or landmark, as this makes it easier to compare.

@@

Since we are using a deformable display, one of the visualizations that makes sense to test is that of 3D volumes, represented using \textit{slices}(textures). This is because these slices are changed based on depth and we can simulate this by pushing an pulling into the display. Further, volume slicing interaction means we can also test how the compensation works when we have a dynamic scene/object. However, an issue with this scenario was that the slices were changing too fast, which, in turn, changed the shape of the object(brain) and made it more difficult for users to judge the compensation effect.

@@

One attempt to solve the problem was to use depth smoothing as described in section \ref{sec:DepthSmoothing}. The result after spatial and temporal averaging was still not perfect - you could still see the difference between the two methods, because of the dynamic noise in the compensated image. However, the smoothing did reduce the overall noise and improved image quality.

@@

Some questions were also asked during the tasks. These are included in the \textit{scenario notes} document, available under appendix \ref{appendix:C}. For the scenarios with the grid, text and columns, feedback was collected regarding all of the three points tested. Similar scenario-specific questions were asked, as mentioned in the document(e.g.: how straight do you think the grid lines appeared with method B?).

@@

However, it is useful to also look in a bit more detail at statistical significance. The output of the t-tests is shown in table \ref{table:POA_Test}. For all non-significant cases, the null hypothesis could not be rejected. Since the confidence interval is 95\%, we would normally need a significance level of $<= 0.05$. However, we have a case of multiple testing, because we have one hypothesis tested for each scenario. As described in \cite{bonferroni}, we need to adjust the parameter $\alpha$, by dividing it with the number of tests(Bonferroni correction). In this case we have $\alpha = 0.05$ and $n$ = 5, which gives $\frac{\alpha}{n} = \frac{0.05}{5} = 0.01$. We must test against this value rather than the original $0.05$. However, in this case, the results are not affected.

@@

Further more, the t-test results are shown in table \ref{table:OPOA_Test}; note that this time we only have one pair. As mentioned before, the missing values were replaced with 0 when running the t-test(the user was forced to select a preference for either method A or B). The results are statistically significant, with a value comfortably below the $0.05$ threshold(there is no need for a correction in this case, since the question is not asked for each scenario, so we only have one test).

@@

Further more, even though the other scenario tests are not statistically significant, the preference of the compensated method was higher for all scenarios, except Text. So, in general, the results indicate that the preference was largely in favor of this method.

@@

Although there was no specific question regarding latency, several users mentioned that the compensation has a "time delay" or that "it takes some time to adjust"(there are 5 such occurrences in total, mentioned by 3 different participants). During the study, the scenarios were run at about 30 fps, because it had the depth smoothing functions enabled. However, this issue was also mentioned by the participant in the second pilot study, when the frame rate was about 60 fps. Therefore, this issue may not be related directly to rendering, although this has not been investigated.

@@

For this point, participants thought that the compensated method was preserving the appearance much  more often than the one which was not compensated(27 vs 3 occurrences). Some examples of items included in this category include: "keeps proportions better", "keeps the size of the object" for the uncompensated method and
"maintains proportions better", "better for the center", "resembles original better", "flattened image", "straighter" for the other. As we can see, the qualitative information is overwhelmingly in favor of the compensated method for this point.

@@

Regarding distortion, there were 19 comments for the compensated method ("distorts a lot", "deforming the image on the side", "the corner is more distorted", "it's going the wrong way - left", "letters go to the left", "moves to the left", "makes columns bend the other way") and only 6 for the uncompensated ("shifting everything", "doesn't compensate", "complete distortion", "perspective deformed"). From these comments we again see that the compensation is too much to the left of the deformed area, and actually, there are 8 such occurrences. This suggests this is the main problem, but one user also shared afterwards that "it was more corrected on the point", even though it looked very distorted. This happens because the setup is not calibrated, and the depth camera is not 100\% orthogonal to the display. This introduces misalignments near the sides if the camera is tilted, even slightly(which means the compensation will take place in a wrong area – with an offset; several users also complained of distortions “underneath the finger”).

@@

For this point, the preservation of appearance category has 14 entries for the compensated method("straighter", "preserves better", "fixing things around the finger", "maintains relative positions"), while only 3 for the uncompensated("more like the left side; keeps proportions better", "more straight" etc.). It is important to note, that although there are many more statements indicating that the compensating method preserves appearance, several of these also include additions like "not as good as center" or "better, but still bent", "not that straight as before(center)".

@@

In the comparison in figure \ref{fig:BrainScanComparison} we can see that the uncompensated method makes the object smaller as we push in, while the compensated one tries to retain its size. However, there may again be some overcompensation, since the image gets a bit larger than the original. A further issue, that may not be apparent in the video or the mentioned figure, is that overcompensation can also happen sideways, and then it seems that the shape of the object is being changed. However, both these issues are present(again) due to calibration issues(there is an offset towards the sides, and the amount of compensation changes with the field of view and distance of the virtual camera, as explained in section \ref{sec:VirtualCameraAlignmentToClothArea}).

@@

As described in section \ref{sec:Empirical Study Results}, we have obtained statistically significant results only for the map scenario and the overall case - for both preservation of appearance and preference. Even though the other scenario tests were not significant, the compensated method almost always scored higher(the only exception being the preference for the Text scenario). The results for the Map had the highest rating (from the scenarios) for the compensated method. This is because the method which was not corrected made the text and roads very small and virtually unreadable, while pushing into the display. On the other hand, the compensated method retained the proportions fairly well.

Even though the consistency tests were not statistically significant, we could see that the results indicated that the behavior of the non-compensated method was more consistent. From the qualitative analysis presented in section \ref{sec:QualitativeAnalysis}, we can obtain more information about what differences participants perceived, and how they differ between the center, left and right points. The analysis indicates that the method compensated best for the center point, and worst for the left point. This issue is due to distortions that are introduced in the image by the compensated method(participants described this as a different kind of distortion), due to an offset compensation. This happens because the system is not calibrated and the components are not perfectly aligned. Although there were issues on the sides of the image, participants still selected the compensated method over the normal one, when asked to rate the methods over all cases. Since, as mentioned before, these tests are statistically significant, it is reasonable to believe that the compensation both preserved the image proportions better and was the preferred method for working with a deformable display(when considering the overall results).

@@

Moreover, an important extension is to provide a useful interface for developing interactive scenarios. Currently, only the brain scan scenario is interactive. Even so, it only checks for the depth of the center point in the image when changing textures, so it is a limited and naive approach. Although interaction is not the focus of this project, it is a necessary extension in order to thoroughly test the distortion correction in dynamic environments. For example, if we want to move a cube around in a 3D scene. In this case, latency becomes very important. It would be interesting to test if the cube would appear undistorted to users in such an environment.

@@

Deformable displays are a newly emerging type of natural user interfaces. One of the issues related to such displays is the distortion that appears from the deformation of the object. The image will follow the shape of the display - it will not look like the intended output on a flat screen; this is called distortion.

@@

The compensation was implemented using a time-of-flight depth sensor, a standard projector and a cloth material(in a wooden frame) to display the images on. The model of the cloth is updated dynamically in a virtual scene(in real-time, up to about 60 fps, when not using noise reduction functions), and we can project any texture we want, to represent our final image contents. The depth camera is set to capture the back side of the display, on which we also project the image. When interacting from the front, pushing into the display is therefore modeled from the back-side, creating an extrusion in the corresponding area(see the thesis video for a demonstration). This enlarges the respective area, which negates the shrinking effect induced by moving that part of the display closer to the projector - this negates the distortion, to a certain extent. The main components have therefore been tested and working, however, the system does have a few limitations, as described in the following paragraph.

@@

However, for a dense set of points, such as that obtained from the depth data, a naive triangle strip triangulation has provided adequate results - the deformation is clearly visible, both in the mesh and on the texture applied to the mesh. For this reason, Delaunay triangulation has not been attempted, although it may improve visual quality. It remains to be determined, as future work, if the computational cost involved is worth the possible increase in visual quality.

@@

Many aspects from a classical implementation of marching cubes can also be used for a GPU implementation. This is the case with the basic look-up tables used to store edges and triangles. A version of these tables used in several implementations is available in an on-line article(see \cite{bourke1994}), where the full source code is also available. Although no version implemented using modern OpenGL was found, an implementation using a geometry shader is described in another on-line article(see \cite{crassin2007}). A geometry shader is simply a shader program that can be used to manipulate geometry directly - it allows the GPU to create geometry. This shader takes a single primitive as input(e.g.: a vertex) and outputs zero or more primitives(e.g.: zero or more triangles). 
This implementation is based on \cite{bourke1994}, with the main marching cubes logic inside the geometry shader. However, this implementation also uses deprecated OpenGL functionality. The main idea is that data can be stored inside 2D and 3D textures and can then be passed to the geometry shader as uniforms:
